---
title: "Inequalities in Access to Neonatal Critical Care"
author: "Ashley Dickson"
date: today
format:
  html:
    toc-location: left
    number-sections: true
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)

library(bit64)
library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)

# NHS colour palette
nhs_blue <- "#005EB8"
nhs_dark_blue <- "#003087"
nhs_light_blue <- "#41B6E6"
nhs_pink <- "#AE2573"
nhs_green <- "#009639"
nhs_grey <- "#768692"

# Load data for summary stats
data <- readRDS("data/ncc_base_dataset.rds")
data <- data %>%
  mutate(Gestation_Weeks = as.numeric(Gestation_Length_At_Birth) / 7)
```

::: {.callout-tip}
## What changed in this revision (June 2026)
Responding to feedback from AE and KR:

- **[A6]** Reconciled the births figures — the primary-model sample (~476,000) is now distinguished from the full cohort (506,656), see [Methods](#phase-1-hierarchical-logistic-regression).
- **[K2]** Added a [Data and Code Availability](#data-and-code-availability) section.

The analytical changes (funnel plot, Black under-admission test, funnel-outlier completeness diagnostic, deprivation-morbidity check, and others) are in the [Findings report](modelling_results.html) and its companion [Technical Appendix](technical_appendix.html). A July 2026 editorial pass tightened both reports and aligned the key messages with the revised findings; a further restructuring split the Phase 1 report into a short Findings document and a separate Technical Appendix, so the findings are readable without the full methodology and diagnostics getting in the way. Nothing was deleted — every number in the Findings report traces back to the appendix.

*(Temporary review scaffolding; removed before final publication.)*
:::

# Overview

Reducing health inequalities is a core priority for the NHS, as set out in the NHS Long Term Plan and the Core20PLUS5 approach. Ensuring equitable access to neonatal critical care (NCC) is essential for giving every baby the best start in life.

This analysis examines whether ethnicity, socioeconomic deprivation, and hospital-level factors are associated with inequalities in NCC access in England, using linked NHS administrative data from the Maternity Services Dataset (MSDS) and Critical Care Minimum Dataset (CCMDS).

The analysis was scoped in two phases, one per question:

1. Who gets admitted? Are some groups less likely to receive NCC after adjusting for clinical need? (Phase 1 — complete, reported here and in the [Findings report](modelling_results.html); full methodology in the [Technical Appendix](technical_appendix.html).)
2. How quickly are they admitted? Among babies with clinical need, do some groups wait longer? (Phase 2 — on hold pending better data; see below.)

::: {.callout-warning}
## Phase 2 status (May 2026)

The planned survival analysis of time from birth to NCC admission is paused. Investigation found that the MSDS table available in this environment (`MSD401BabyDemographics_1`) records year and month of birth but no day of birth — the field previously assumed to encode day-of-month is in fact an AM/PM meridian flag. As a result, any "hours from birth" figures that were derived in earlier drafts of this work are inaccurate and have been withdrawn.

A real day-of-birth field is available in a separate data environment, which is being applied for. Phase 2 will resume once that access is in place. Phase 1 (admitted yes/no) is unaffected — it does not depend on birth datetime.
:::

# Study Population

```{r study_pop, echo=FALSE}
n_births <- nrow(data)
n_ncc <- sum(data$NCC_Admitted, na.rm = TRUE)
ncc_rate <- round(100 * mean(data$NCC_Admitted, na.rm = TRUE), 1)
```

The analysis covers `r comma(n_births)` hospital births in England during financial year 2023/24, of which `r comma(n_ncc)` (`r ncc_rate`%) were admitted to neonatal critical care, combining two admission sources (MSDS and the Critical Care Minimum Dataset — see the [Technical Appendix](technical_appendix.html#outcome-definition) for how the combined flag is derived and why the two sources disagree).

# Methods

## Phase 1: Hierarchical Logistic Regression

We fitted mixed-effects logistic regression models with random intercepts for delivery trust, asking whether the probability of NCC admission varies by ethnicity and deprivation after adjusting for clinical need. (See the Phase 1 Methods section for why we cluster at trust rather than site level.)

- Primary model: Adjusts for gestation, ethnicity (9 disaggregated groups), and IMD quintile, with trust random intercepts (~476,000 births — the full cohort of `r comma(n_births)` less records missing gestation, ethnicity or IMD)
- Sensitivity model: Additionally adjusts for birth weight and Apgar 5-minute score (~219,000 births with complete clinical data)

This two-model design lets the primary model run on nearly the whole cohort, while the sensitivity model checks the findings hold with fuller clinical adjustment. Ethnicity is coded using disaggregated NHS categories (Indian, Pakistani and Bangladeshi separately, rather than a single "Asian" group); the results bear this choice out, since these groups differ materially.

Both are mixed-effects (multilevel) models: each trust gets its own baseline admission rate, and the ethnicity and deprivation effects are estimated *within* trusts. This matters because baseline NCC admission rates vary widely between hospitals — through clinical thresholds, capacity and reporting practice. Without the hospital baseline in the model, a group of babies concentrated in high-admitting hospitals would look over-admitted when the effect is really the hospital's, not the group's. The estimates therefore answer a like-for-like question: *for two clinically similar babies born in the same trust, does the chance of admission differ by ethnicity or deprivation?*

Full results: [Phase 1: Findings](modelling_results.html) — full methodology and diagnostics: [Phase 1: Technical Appendix](technical_appendix.html)

## Phase 2: Cox Proportional Hazards with Shared Frailty *(on hold)*

The planned Phase 2 analysis would fit Cox models with a shared frailty term for delivery trust — the survival analysis analogue of the mixed-effects logistic model — to isolate the timing/delay question from the access question addressed in Phase 1.

This phase is currently paused: the MSDS extract available in this environment does not contain a true day-of-birth field, so hours-from-birth cannot be computed reliably. Once the alternative data environment is in place, the planned specification is:

- Primary model: Adjust for gestation, ethnicity, IMD, and delivery unit type (NICU/LNU/SCBU), with hospital frailty
- Sensitivity model: Additionally adjust for birth weight and Apgar score
- Analysis restricted to births at hospitals with a recognised NNAP neonatal unit

# Key Findings

Phase 1 finds a system that looks broadly need-led: gestation dominates the model, Black babies sit at parity with White British despite higher average clinical need, and the deprivation gradient tracks a genuine morbidity gradient rather than over-admission. Two ethnic groups (Pakistani, Chinese) and one large residual signal (between-trust variation, ICC ~32%, substantially a recording-definition artefact rather than an admission-practice one) warrant follow-up. The full key messages, with the evidence behind each, are in the [Findings report](modelling_results.html#key-messages); Phase 2 (the timing question) remains on hold pending data access (see above).

# Data Sources

The analysis links three national NHS datasets, accessed via the NHS England Unified Data Access Layer (UDAL):

| Source | Content | Role |
|---|---|---|
| MSDS (Maternity Services Dataset) | Birth details, baby and maternal demographics, pregnancy booking, neonatal admissions | Base cohort of births |
| CCMDS (Critical Care Minimum Dataset, via PbR_CC_Monthly) | Daily-level NCC activity, organ support, care levels | NCC admission status and timing |
| APCS (Admitted Patient Care Spells) | Hospital spell data with NHS Numbers | Linkage between MSDS and CC data |

Records are linked via pseudonymised NHS Number, with a 90-day date proximity filter when linking births to CC admissions. The dataset covers FY 2023/24. NNAP unit data from the [RCPCH 2024 report](https://www.rcpch.ac.uk/resources/NNAP-summary-report-2024-data) is used to classify delivery sites by unit type.

# Data and Code Availability

All analysis code is version-controlled and published on GitHub at [AshleyDicksonNHS/ncc_inequalities_analysis](https://github.com/AshleyDicksonNHS/ncc_inequalities_analysis); the rendered reports are served from that repository via GitHub Pages.

No patient-level data is held in the repository — the `.gitignore` excludes all data files. The extracted analytical dataset (pseudonymised, record-level) is stored separately in a restricted NHS England SharePoint location for long-term NHSE/DHSC access, consistent with information-governance requirements. The published HTML reports contain only aggregate results.

---

Report generated: `r Sys.time()`

R version: `r R.version.string`
