---
title: "Inequalities in Access to Neonatal Critical Care: Phase 1 Findings"
author: "Ashley Dickson"
date: today
format:
  html:
    toc-location: left
    number-sections: true
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)

library(bit64)
library(tidyverse)
library(lme4)
library(knitr)
library(kableExtra)
library(scales)

# NHS colour palette
nhs_blue <- "#005EB8"
nhs_pink <- "#AE2573"
nhs_aqua <- "#00A9CE"

# Load models
model2 <- readRDS("data/model2_mixed_effects.rds")
model3 <- readRDS("data/model3_sensitivity.rds")
```

::: {.callout-tip}
## What changed in this revision (June–July 2026)
This draft responds to feedback from AE and KR. Key changes:

- [K3] Clarified that the cohort is babies born in 2023/24, not continuing-stay.
- [K5a] Rewrote variable selection around causal role (confounder / exposure / mediator), with a full candidate-variable table and the case against a data-driven (stepwise/LASSO) search for a causal access question.
- [A1] Rewrote the Chinese-group key message in plain language.
- [A3a/A3b] Reframed the deprivation finding around the unmeasured-need vs genuine-access question, with a direct morbidity-by-deprivation check.
- [A4/A4b] New funnel plot of case-mix-adjusted admission rates, with a completeness diagnostic showing the outlier signal is a recording-definition issue, not an admission-practice one.
- [A2] New test for trusts where Black admission is significantly lower than White.
- [K4] New broad MBRRACE-comparable ethnicity grouping presented alongside the detailed one.
- [A5c] New exploratory ethnicity × deprivation interaction: no evidence of a steeper deprivation penalty for minority groups.
- [A5a] New length-of-stay analysis: the crude Black excess is entirely case-mix; Pakistani and Bangladeshi babies stay longer than expected.
- [A5b] New descriptive mortality section: the Pakistani excess is not explained by gestational mix; the Black excess mostly is.
- Restructuring (July 2026): this report was split into a short Findings report (this document) and a new [Technical Appendix](technical_appendix.html) containing full methodology, model specifications, and diagnostic detail. This was done because the combined report had grown too long; splitting it keeps the findings readable without losing any of the underlying rigor.

*(This box is for temporary review and will be removed before final publication.)*
:::

# Executive Summary

This report examines whether, after accounting for clinical need, access to neonatal critical care (NCC) differs by ethnicity or deprivation, using around 476,000 English hospital births in 2023/24. The analysis uses hierarchical logistic regression: a primary model adjusting for gestation, a sensitivity model additionally adjusting for birth weight and Apgar score, and a baseline admission rate for each delivery trust. Full methodology and every diagnostic check behind these numbers are in the [Technical Appendix](technical_appendix.html).

## Key Messages

1. Clinical need drives admission. Each additional week of gestation reduces the odds of NCC admission by around 38%; ethnicity and deprivation effects are small by comparison. On its central question, the system appears need-led.

2. There is no evidence of under-admission for Black babies at national level. Odds ratios are approximately 1.04 in both the primary and sensitivity models — essentially parity with White British, despite higher average clinical need. This is a finding about admission rates only, not about the wider quality of care received.

3. Two groups warrant follow-up. Pakistani babies' lower odds in the primary model move to near-parity once birth weight and Apgar are added, so are largely explained by clinical case-mix — and the [length-of-stay and mortality](technical_appendix.html#length-of-stay) analyses show the same pattern, consistent with these babies being admitted only once more seriously unwell. Chinese babies show around 24% lower odds even after adjustment; three broad explanations are consistent with this and the data cannot separate them (see [Ethnicity](#ethnicity)).

4. The deprivation gradient is consistent with need rather than over-admission. Odds are around 9% higher for the most deprived quintile relative to the least, stable after clinical adjustment, and a [direct morbidity check](technical_appendix.html#imd-morbidity) shows clinical need itself rising with deprivation in ways the model does not capture.

5. Between-trust variation is the largest residual signal (ICC around 32%), but it bundles admission practice with recording practice and cannot be read at face value. The [funnel-outlier diagnostic](technical_appendix.html#outlier-completeness) shows the most extreme trusts record NCC for 21–40% of their term babies, against roughly 7% nationally — a recording signature, not an admission-practice one. Standardising what counts as an NCC day in PbR-CC returns is the highest-value data improvement identified by this work.

6. These results address whether a baby is admitted, not when. The planned Phase 2 analysis of admission timing is on hold pending access to a data environment with a usable date of birth.

::: {.callout-note}
## Terms used in this report
- **Odds ratio (OR)**: how much more, or less, likely a group is to be admitted than the reference group, after adjustment. OR 1.00 = parity; OR 1.10 = 10% higher odds; OR 0.90 = 10% lower odds.
- **ICC (intraclass correlation)**: the share of variation in admission that sits between hospitals rather than between babies.
- **BH-adjusted p-value**: a p-value corrected for testing several groups at once, so that chance findings among (for example) 9 ethnicity groups aren't over-interpreted.
- **O/E ratio**: observed cases (deaths or admissions) divided by the number expected given a group's clinical profile. Above 1 means more than expected; below 1 means fewer.
- **Rate ratio (RR)**: the same idea as an odds ratio, applied to a count outcome (e.g. days of care) rather than a yes/no one.
:::

---

# Methods, in brief

```{r study_pop, echo=FALSE}
n_primary <- nobs(model2)
n_sensitivity <- nobs(model3)
```

The primary model is a hierarchical (mixed-effects) logistic regression — `NCC_Admitted ~ Gestation + Ethnicity + IMD`, with a random intercept per delivery trust — fit on `r format(n_primary, big.mark = ",")` hospital births in England in FY 2023/24 with complete data on these fields. A sensitivity model adds birth weight and Apgar 5-minute score, on the `r format(n_sensitivity, big.mark = ",")` births (`r round(100*n_sensitivity/n_primary, 1)`%) where these are also complete.

The random intercept means every comparison is effectively *within* the same hospital: the estimates answer, for two clinically similar babies born at the same trust, whether the chance of admission differs by ethnicity or deprivation. Ethnicity is coded using nine disaggregated NHS categories rather than broad groups, because — as the results below show — broad grouping conceals real differences between, for example, Pakistani, Indian and Bangladeshi babies. Deprivation uses 2015-vintage IMD quintiles of the mother's recorded address.

Full methodology — the causal reasoning behind variable selection, cohort filters, model equations, and why clustering is at trust rather than site level — is in the [Technical Appendix](technical_appendix.html#methods).

---

# Results

## Clinical Factors

```{r model2_fe, echo=FALSE}
fe <- fixef(model2)
se <- sqrt(diag(vcov(model2)))
z_val <- fe / se
p_val <- 2 * pnorm(-abs(z_val))

model2_or <- data.frame(
  term = names(fe),
  OR = exp(fe),
  CI_lower = exp(fe - 1.96 * se),
  CI_upper = exp(fe + 1.96 * se),
  p.value = p_val,
  raw_term = names(fe),
  stringsAsFactors = FALSE
) %>%
  mutate(
    family = case_when(
      str_detect(raw_term, "Ethnicity") ~ "ethnicity",
      str_detect(raw_term, "IMD")       ~ "imd",
      TRUE                               ~ "other"
    )
  ) %>%
  mutate(
    term = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "Gestation_Weeks" ~ "Gestation (per week)",
      str_detect(term, "Ethnicity_Baby_Grouped") ~ str_replace(term, "Ethnicity_Baby_Grouped", "Ethnicity: "),
      str_detect(term, "IMD_Quintile") ~ str_replace(term, "IMD_Quintile", "IMD: "),
      TRUE ~ term
    )
  )

var_trust <- as.numeric(VarCorr(model2)$Delivery_Trust)
icc <- var_trust / (var_trust + (pi^2 / 3))
```

Gestation is by far the strongest predictor of NCC admission: each additional week reduces the odds by around `r round((1 - exp(unname(fe["Gestation_Weeks"]))) * 100, 0)`%. This confirms the model is capturing genuine clinical need, and it is the backdrop against which the ethnicity and deprivation effects below — much smaller in comparison — should be read. Full fixed-effects table in the [Technical Appendix](technical_appendix.html#model-2-mixed-effects-logistic-regression).

```{r forest_plot, echo=FALSE, fig.height=6}
#| label: fig-forest
#| fig-cap: "Odds ratios for NCC admission by ethnicity and deprivation (adjusted for gestation, with hospital random effects)"

plot_data <- model2_or %>%
  filter(!term %in% c("Intercept", "Gestation (per week)")) %>%
  mutate(
    category = case_when(
      str_detect(term, "Ethnicity") ~ "Ethnicity",
      str_detect(term, "IMD") ~ "Deprivation"
    ),
    term = str_replace(term, "Ethnicity: ", ""),
    term = str_replace(term, "IMD: ", ""),
    term = factor(term, levels = rev(term))
  )

ggplot(plot_data, aes(x = OR, y = term, color = category)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper), height = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Ethnicity" = nhs_blue, "Deprivation" = nhs_pink)) +
  scale_x_continuous(breaks = seq(0.7, 1.3, 0.1)) +
  facet_grid(category ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Adjusted Odds Ratios for NCC Admission",
    subtitle = "Reference groups: White British (ethnicity), Least Deprived quintile (IMD)",
    x = "Odds Ratio (95% CI)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Ethnicity {#ethnicity}

Most ethnic groups sit within a few percent of parity with White British babies. The Black group sits at near-exact parity; Pakistani and Chinese babies sit below parity; Indian and Bangladeshi babies sit slightly above. How each estimate responds when birth weight and Apgar are added (the sensitivity model) is the key to interpreting them:

| Group | Primary OR | Sensitivity OR | Interpretation |
|-------|-----------|----------------|----------------|
| Pakistani | 0.89 | 0.97 | Moves towards parity — clinical factors explain much of the difference |
| Chinese | 0.70 | 0.76 | Still notably lower, at the boundary of nominal significance in the smaller sample |
| Indian | 1.05 | 1.16 | Increases further — higher access after clinical adjustment |
| Bangladeshi | 1.06 | 1.17 | Similar pattern to Indian |
| Black | 1.04 | 1.04 | Stable — consistent with equitable access |

The differences among Indian, Pakistani and Bangladeshi babies after adjustment are the clearest demonstration of why this report uses disaggregated ethnicity coding, rather than the broader groups MBRRACE-style reporting typically uses — a broader grouping gives a reassuring headline (no broad group far from parity) but conceals this spread and the Chinese finding below (full comparison in the [Technical Appendix](technical_appendix.html#broad-ethnicity)).

Black babies sit at national parity (OR 1.04, stable after adjustment) despite carrying higher average clinical need on average — no evidence of national under-admission. One of the 92 trusts with enough Black births to test showed a significantly lower rate and has been flagged for a case-mix-standardised follow-up (full test in the [Technical Appendix](technical_appendix.html#black-underaccess-screen)).

Chinese babies show the largest unexplained gap: around 24% lower adjusted odds. Three broad explanations are consistent with this and the data cannot separate them — unmeasured clinical need, a smaller and less stable sample, or data/access factors (differences in care-seeking, or possible under-recording). This is a question for follow-up, not a conclusion (see [Recommendations](#recommendations) and the [Technical Appendix](technical_appendix.html#ethnicity) for detail).

## Deprivation

```{r imd_interpretation, echo=FALSE}
imd_or <- model2_or %>%
  filter(str_detect(term, "IMD")) %>%
  arrange(desc(term))

imd_summary <- imd_or %>%
  mutate(
    pct_diff = round((OR - 1) * 100, 0),
    direction = ifelse(OR > 1, "higher", "lower")
  )
```

Babies from the most deprived quintile have `r abs(imd_summary$pct_diff[1])`% `r imd_summary$direction[1]` odds of NCC admission than the least deprived, stepping up consistently across quintiles. This could mean the model is picking up genuine clinical need that correlates with deprivation but isn't fully captured by gestation, birth weight and Apgar alone — or it could mean a genuine difference in admission threshold. The two are hard to tell apart from the model alone.

A direct check of morbidity by deprivation quintile supports the first reading: every marker of clinical need (preterm birth, very preterm birth, low birth weight) rises monotonically with deprivation, and low birth weight among *term* babies alone very nearly doubles from least to most deprived (full quintile table in the [Technical Appendix](technical_appendix.html#imd-morbidity)). We also checked whether deprivation carries a steeper penalty for any one ethnic group — it doesn't appear to, though the data for smaller ethnic groups are too thin to be confident either way (technical detail in the [Appendix](technical_appendix.html#eth-imd-interaction)).

## Hospital Variation

Once gestation, ethnicity and deprivation are accounted for, the trust a baby is born in is the largest remaining source of variation in the model — an ICC of around `r round(icc*100, 0)`%. This bundles together genuine differences in clinical practice, referral patterns, and how providers record NCC activity, and should not be read as a clean measure of admission practice alone (full breakdown of what it can and can't mean in the [Technical Appendix](technical_appendix.html#hospital-variation)).

The funnel plot below checks each trust's case-mix-adjusted admission rate against control limits; a small number of trusts sit well outside them.

```{r funnel_plot, echo=FALSE, fig.height=6}
#| label: fig-funnel
#| fig-cap: "Funnel plot of case-mix-adjusted NCC admission rates by delivery trust. Points outside the 99.8% limits admit at a rate that case-mix and chance alone do not explain. Adjusted for gestation, ethnicity and IMD via the primary model's fixed effects (trust random effect removed)."

mf <- model2@frame
mf$.observed <- as.numeric(mf$NCC_Admitted)
mf$.expected <- predict(model2, type = "response", re.form = NA)

funnel_df <- mf %>%
  group_by(Delivery_Trust) %>%
  summarise(n = dplyr::n(),
            O = sum(.observed),
            E = sum(.expected),
            .groups = "drop") %>%
  mutate(Delivery_Trust = as.character(Delivery_Trust)) %>%
  filter(n >= 100)

scale_factor <- sum(funnel_df$O) / sum(funnel_df$E)
p_bar <- sum(funnel_df$O) / sum(funnel_df$n)

funnel_df <- funnel_df %>%
  mutate(E   = E * scale_factor,
         SAR = O / E,
         adj_rate = 100 * p_bar * SAR)

z95  <- qnorm(0.975)
z998 <- qnorm(0.999)

pearson <- (funnel_df$O - funnel_df$E) / sqrt(funnel_df$E)
zq      <- quantile(pearson, c(0.1, 0.9), na.rm = TRUE)
z_wins  <- pmin(pmax(pearson, zq[1]), zq[2])
phi     <- max(1, sum(z_wins^2) / length(z_wins))
infl    <- sqrt(phi)

limit_grid <- tibble(E = seq(min(funnel_df$E), max(funnel_df$E), length.out = 400)) %>%
  mutate(u95  = 100 * p_bar * (1 + z95  * infl / sqrt(E)),
         l95  = 100 * p_bar * pmax(0, 1 - z95  * infl / sqrt(E)),
         u998 = 100 * p_bar * (1 + z998 * infl / sqrt(E)),
         l998 = 100 * p_bar * pmax(0, 1 - z998 * infl / sqrt(E)))

funnel_df <- funnel_df %>%
  mutate(outlier = adj_rate > 100 * p_bar * (1 + z998 * infl / sqrt(E)) |
                   adj_rate < 100 * p_bar * (1 - z998 * infl / sqrt(E)),
         Direction = case_when(!outlier ~ "Within limits",
                               SAR > 1   ~ "Higher than expected",
                               TRUE      ~ "Lower than expected"))

y_top <- min(100, max(funnel_df$adj_rate, na.rm = TRUE) * 1.15)

ggplot() +
  geom_line(data = limit_grid, aes(E, u95),  linetype = "dashed", colour = "grey55") +
  geom_line(data = limit_grid, aes(E, l95),  linetype = "dashed", colour = "grey55") +
  geom_line(data = limit_grid, aes(E, u998), linetype = "dotted", colour = "grey30") +
  geom_line(data = limit_grid, aes(E, l998), linetype = "dotted", colour = "grey30") +
  geom_hline(yintercept = 100 * p_bar, colour = nhs_blue, linewidth = 0.6) +
  geom_point(data = funnel_df, aes(E, adj_rate, colour = Direction, size = n), alpha = 0.75) +
  scale_colour_manual(values = c("Within limits" = "grey65",
                                 "Higher than expected" = nhs_pink,
                                 "Lower than expected" = nhs_aqua)) +
  scale_size_continuous(range = c(0.7, 4), guide = "none") +
  labs(title = "Funnel plot: case-mix-adjusted NCC admission rates by trust",
       subtitle = paste0("Overdispersion-adjusted limits (dashed 95%, dotted 99.8%); blue line = national rate (",
                          round(100 * p_bar, 1), "%)"),
       x = "Expected NCC admissions (case-mix adjusted)",
       y = "Adjusted NCC admission rate (%)",
       colour = NULL) +
  coord_cartesian(ylim = c(0, y_top)) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom")
```

Looking closely at the trusts outside the limits, the evidence points to a recording issue rather than a practice one: they record NCC for 21–40% of *term* babies, against ~7% nationally, most likely because they count a broader range of neonatal unit activity as "NCC" days in their PbR-CC returns than their peers do (full diagnostic, which rules out missing records and missing fields as alternative explanations, in the [Technical Appendix](technical_appendix.html#outlier-completeness)). Standardising what counts as an NCC day is the single highest-value data improvement this analysis points to (Recommendation 1).

## Length of Stay and Mortality

```{r extended_setup, echo=FALSE}
a5a <- readRDS("data/a5a_los_summary.rds")
rr <- a5a$rr_table %>% filter(raw_term != "(Intercept)")
rr_of <- function(pattern) rr %>% filter(str_detect(raw_term, fixed(pattern)))
rr_black <- rr_of("Black")
rr_pak   <- rr_of("Pakistani")
fmt_rr <- function(r) sprintf("%.2f (%.2f–%.2f)", r$RR, r$Lower, r$Upper)

a5b <- readRDS("data/a5b_mortality_summary.rds")
oe_of <- function(tbl, pattern) tbl %>% filter(str_detect(as.character(Group), fixed(pattern)))
oe_pak <- oe_of(a5b$oe_eth, "Pakistani")
fmt_oe <- function(o) sprintf("%.2f (%.2f–%.2f)", o$OE, o$lo, o$hi)
```

Two further checks ask not just who is admitted, but what happens next.

**Length of stay.** Among admitted babies, the crude excess stay length for Black babies disappears entirely once adjusted for gestation (RR `r fmt_rr(rr_black)`) — babies are admitted at parity *and* receive the expected amount of care once admitted. Pakistani babies stay around 9% longer than expected for their gestation (RR `r fmt_rr(rr_pak)`), alongside their lower admission odds — a pattern consistent with a higher effective admission threshold, though this data cannot fully separate that from greater within-gestation morbidity (full model in the [Technical Appendix](technical_appendix.html#length-of-stay)).

**Mortality.** A descriptive, dual-source mortality flag — not a substitute for MBRRACE's linked-registration data, and read with heavy ascertainment caveats — shows the Black crude excess is mostly explained by gestational mix, but the Pakistani excess is not: `r round(oe_pak$OE * 100 - 100)`% more recorded deaths than the group's gestational profile predicts. Combined with the admission and length-of-stay findings, this points to the same follow-up: severity at the point of admission (Recommendation 6). Full detail, including why this flag is not a substitute for national mortality statistics, is in the [Technical Appendix](technical_appendix.html#mortality).

---

# Recommendations {#recommendations}

The analysis is reassuring on its headline question — no signal of national under-admission for Black babies — but it surfaces concrete follow-ups. In rough priority order:

1. Standardise what counts as an "NCC day" in PbR-CC returns. The [funnel diagnostic](technical_appendix.html#outlier-completeness) shows the most extreme provider variation is recording breadth, not admission practice: outlier trusts record NCC for 21–40% of their term babies against ~7% nationally. A tighter definition — or an audit of the outlier trusts' returns — would sharpen every downstream use of this data, including the ICC and any future benchmarking.

2. Improve completeness of birth weight and Apgar in MSDS. Jointly complete for under half the cohort, this is the single biggest constraint on case-mix adjustment — it is why the primary model must lean on gestation alone.

3. Follow up the one trust flagged for lower Black admission (identity held internally) with a case-mix-standardised check — a per-trust Black observed/expected ratio — before drawing any conclusion about access.

4. Investigate the Chinese-group gap. Around 24% lower adjusted odds with three candidate explanations this data cannot separate; richer clinical covariates or direct provider engagement would be needed to distinguish unmeasured need from sample-size instability or from access/recording factors.

5. Extend to outcomes and timing. Length of stay and a descriptive mortality look are analysed in the [Technical Appendix](technical_appendix.html) ([stay length](technical_appendix.html#length-of-stay), [mortality](technical_appendix.html#mortality)) — both add Pakistani findings that sharpen the case for the Phase 2 timing analysis once the day-of-birth data environment is available. A definitive mortality analysis needs linked death registrations (MBRRACE territory), not this extract.

6. Follow up the Pakistani stay-length excess. Fewer admissions plus longer stays once admitted is consistent with a higher effective admission threshold, but equally with greater within-gestation morbidity; distinguishing them needs severity-at-admission data (care level at day 1, organ support) — a planned extension.

# Limitations

1. Clinical covariate completeness: Birth weight is recorded for ~46% of births and Apgar for ~79%; jointly they limit the sensitivity model to under half the cohort. The primary model uses gestation alone to preserve sample size.

2. Single year of data: Results may not generalise to other time periods.

3. Residual confounding: While ethnicity is now disaggregated (e.g., Indian, Pakistani, Bangladeshi separately), within-group heterogeneity remains. Black ethnicities are still grouped together in the model.

4. Selection bias: Analysis restricted to births with complete data on gestation, ethnicity, IMD and delivery trust — though the retained share is high and similar across providers (see the [funnel diagnostic](technical_appendix.html#outlier-completeness)).

5. Outcome definition: "NCC admission" is whatever providers record as NCC in two administrative sources, and the funnel diagnostic shows the breadth of that recording varies materially between providers. The outcome is best read as "received care recorded as NCC", not "occupied a critical care cot".

---

Full technical detail — model specifications, causal reasoning behind variable selection, and every diagnostic check behind the numbers above — is in the [Technical Appendix](technical_appendix.html).

Report generated: `r Sys.time()`

R version: `r R.version.string`
