---
title: "Inequalities in NCC Access: Survival Analysis Results"
author: "Ashley Dickson"
date: today
format:
  html:
    toc-location: left
    number-sections: true
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)

library(bit64)
library(tidyverse)
library(survival)
library(coxme)
library(knitr)
library(kableExtra)
library(scales)

# NHS colour palette
nhs_blue <- "#005EB8"
nhs_dark_blue <- "#003087"
nhs_light_blue <- "#41B6E6"
nhs_pink <- "#AE2573"
nhs_green <- "#009639"
nhs_grey <- "#768692"
nhs_orange <- "#ED8B00"

# Load models
model6 <- readRDS("data/model6_cox_conditional.rds")
model7 <- readRDS("data/model7_cox_conditional_sensitivity.rds")

# Load survival dataset
surv_data <- readRDS("data/surv_data.rds")
```

# Executive Summary

This report presents results from Phase 2 of the NCC access inequalities analysis: survival modelling of the time from birth to NCC admission. Where the [Phase 1 logistic models](modelling_results.html) asked *whether* a baby is admitted to NCC, this analysis asks *how quickly* — and specifically, whether some groups experience delayed access.

The key innovation is restricting the analysis to babies who were admitted to NCC (i.e., babies we know needed critical care). This isolates the pure timing/delay question from the access question already answered by Phase 1.

```{r exec_setup, echo=FALSE}
fe6 <- fixef(model6)
se6 <- sqrt(diag(as.matrix(vcov(model6))))
fe7 <- fixef(model7)
se7 <- sqrt(diag(as.matrix(vcov(model7))))

# NCC-admitted babies at NNAP trusts
admitted_nnap <- surv_data %>%
  filter(event == 1, At_NNAP_Trust == TRUE,
         !is.na(Ethnicity_Baby_Grouped), !is.na(Gestation_Weeks),
         !is.na(IMD_Quintile), !is.na(Delivery_Unit_Type),
         !is.na(Time_To_NCC_Hours), Time_To_NCC_Hours > 0)
```

Key findings:

- No systematic ethnic delay: Among babies admitted to NCC, all ethnic groups are admitted at similar speed after clinical adjustment. No group shows a statistically significant delay relative to White British babies.
- No deprivation gradient: Deprivation has no effect on admission timing — all IMD quintiles show HR ≈ 1.0.
- Unit type matters: Babies born at LNU sites are admitted ~3% faster than those at NICUs (HR 1.03, p=0.008), likely reflecting the simpler local admission pathway.
- Hospital variation is negligible: The frailty variance dropped to near-zero (0.00003), compared with an ICC of ~32% in the logistic model. Hospitals differ in *who* they admit, not *how fast*.

::: {.callout-important}
## Key Messages

1. Once the decision to admit is made, NCC access is equitable. Ethnicity and deprivation do not delay admission. Clinicians' assertion that care is guided by signs and symptoms — not demographics — is supported by the data.

2. The inequality sits upstream. Phase 1 showed that hospital variation and modest ethnicity effects influence *whether* a baby enters NCC. Phase 2 shows that *once admitted*, the process is fair. The intervention point is the admission threshold, not the admission process.

3. Unit designation shapes the pathway. LNU sites admit slightly faster than NICUs — consistent with local, same-hospital admissions versus the more complex transfers and stabilisation involved at tertiary centres.
:::

---

# Methods

## Study Population

```{r study_pop, echo=FALSE}
n_m6 <- nrow(admitted_nnap %>%
  filter(!is.na(Delivery_Site_Code), Time_To_NCC_Hours <= 672))
n_m7 <- nrow(admitted_nnap %>%
  filter(!is.na(Delivery_Site_Code), !is.na(Birth_Weight_kg),
         !is.na(Apgar_5_Clean), Time_To_NCC_Hours <= 672))
```

The analysis includes babies born in England during FY 2023/24 who were admitted to NCC at a hospital within a recognised NNAP neonatal unit. The NNAP filter restricts to `r comma(n_distinct(admitted_nnap$Delivery_Site_Code))` delivery sites across 105 NHS trusts with designated neonatal units (NICU, LNU, or SCBU), as identified from the [NNAP 2024 unit data](https://www.rcpch.ac.uk/resources/NNAP-summary-report-2024-data).

- Primary analysis (Model 6): `r comma(n_m6)` NCC admissions with complete data on gestation, ethnicity, IMD, and delivery unit type
- Sensitivity analysis (Model 7): `r comma(n_m7)` admissions with additional birth weight and Apgar data

## Outcome

Time from birth to NCC admission (hours). All babies in this analysis have the event — the outcome is *when*, not *whether*. Times are capped at 28 days (672 hours); admissions beyond this are treated as data linkage artefacts. Times ≤0 hours are set to 0.1 (6 minutes) to represent same-hour admission.

## Statistical Approach

We fitted Cox proportional hazards models with a shared frailty term (random intercept) for delivery site — the survival analysis analogue of the mixed-effects logistic model from Phase 1.

Model 6 (Primary):
$$h(t) = h_0(t) \cdot \exp(\beta_1 \text{Gestation} + \beta_2 \text{Ethnicity} + \beta_3 \text{IMD} + \beta_4 \text{UnitType} + w_j)$$

Model 7 (Sensitivity):
$$h(t) = h_0(t) \cdot \exp(\beta_1 \text{Gestation} + \beta_2 \text{BirthWeight} + \beta_3 \text{Apgar} + \beta_4 \text{Ethnicity} + \beta_5 \text{IMD} + \beta_6 \text{UnitType} + w_j)$$

where $w_j \sim N(0, \sigma^2_w)$ is the shared frailty for delivery site $j$, and $h_0(t)$ is the baseline hazard.

::: {.callout-note}
## Interpreting Hazard Ratios

HR > 1 means faster admission (event occurs sooner). HR < 1 means slower admission (potential delay). This is the opposite intuition to odds ratios — a higher hazard ratio is the "better" outcome here.
:::

## Unit Type Classification

Each delivery site is classified by its highest neonatal unit designation:

- NICU (Level 3): Neonatal Intensive Care Unit — provides the full range of medical and surgical neonatal care
- LNU (Level 2): Local Neonatal Unit — provides high dependency and some short-term intensive care
- SCBU (Level 1): Special Care Baby Unit — provides special care for the least unwell babies

Unit type is included as a fixed effect (reference: NICU). The hospital-level frailty captures residual variation *after* accounting for unit type.

---

# Descriptive Results

## Time to NCC Admission

```{r time_descriptive, echo=FALSE}
ncc_times <- surv_data %>%
  filter(event == 1, At_NNAP_Trust == TRUE, !is.na(Time_To_NCC_Hours),
         Time_To_NCC_Hours > 0, Time_To_NCC_Hours <= 672)

time_summary <- tibble(
  Statistic = c("N", "Median (hours)", "Mean (hours)", "IQR",
                "Within 1 hour", "Within 6 hours", "Within 24 hours", "Within 72 hours"),
  Value = c(
    comma(nrow(ncc_times)),
    round(median(ncc_times$Time_To_NCC_Hours), 1),
    round(mean(ncc_times$Time_To_NCC_Hours), 1),
    paste0(round(quantile(ncc_times$Time_To_NCC_Hours, 0.25), 1), " – ",
           round(quantile(ncc_times$Time_To_NCC_Hours, 0.75), 1)),
    paste0(round(100 * mean(ncc_times$Time_To_NCC_Hours <= 1), 1), "%"),
    paste0(round(100 * mean(ncc_times$Time_To_NCC_Hours <= 6), 1), "%"),
    paste0(round(100 * mean(ncc_times$Time_To_NCC_Hours <= 24), 1), "%"),
    paste0(round(100 * mean(ncc_times$Time_To_NCC_Hours <= 72), 1), "%")
  )
)

time_summary %>%
  kable(caption = "Time from birth to NCC admission (admitted babies at NNAP units)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## Time to NCC by Ethnicity

```{r time_by_ethnicity, echo=FALSE}
eth_times <- ncc_times %>%
  filter(!is.na(Ethnicity_Baby_Grouped)) %>%
  group_by(Ethnicity_Baby_Grouped) %>%
  summarise(
    N = n(),
    `Median (h)` = round(median(Time_To_NCC_Hours), 2),
    `Q25 (h)` = round(quantile(Time_To_NCC_Hours, 0.25), 2),
    `Q75 (h)` = round(quantile(Time_To_NCC_Hours, 0.75), 2),
    `≤1h (%)` = round(100 * mean(Time_To_NCC_Hours <= 1), 1),
    .groups = "drop"
  )

eth_times %>%
  kable(caption = "Time to NCC admission by ethnicity (admitted babies at NNAP units)",
        format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## Time to NCC by Unit Type

```{r time_by_unit, echo=FALSE}
unit_times <- ncc_times %>%
  filter(!is.na(Delivery_Unit_Type)) %>%
  group_by(Delivery_Unit_Type) %>%
  summarise(
    N = n(),
    `Median (h)` = round(median(Time_To_NCC_Hours), 2),
    `Q25 (h)` = round(quantile(Time_To_NCC_Hours, 0.25), 2),
    `Q75 (h)` = round(quantile(Time_To_NCC_Hours, 0.75), 2),
    `≤1h (%)` = round(100 * mean(Time_To_NCC_Hours <= 1), 1),
    `≤6h (%)` = round(100 * mean(Time_To_NCC_Hours <= 6), 1),
    .groups = "drop"
  )

unit_times %>%
  kable(caption = "Time to NCC admission by delivery site unit type") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

---

# Kaplan-Meier Curves

## Broad Ethnicity Groups

Among admitted babies, the Kaplan-Meier curves show the proportion still awaiting NCC admission at each time point. A lower curve means faster access.

```{r km_broad, echo=FALSE, fig.height=7, fig.width=10}
km_clean <- surv_data %>%
  filter(event == 1, At_NNAP_Trust == TRUE,
         !is.na(Ethnicity_Baby_Grouped), !is.na(Time_To_NCC_Hours),
         Time_To_NCC_Hours > 0, Time_To_NCC_Hours <= 72) %>%
  mutate(
    Ethnicity_Broad = case_when(
      Ethnicity_Baby_Grouped %in% c("White British", "White Other") ~ "White",
      Ethnicity_Baby_Grouped %in% c("Indian", "Pakistani", "Bangladeshi",
                                     "Other Asian", "Chinese") ~ "Asian",
      Ethnicity_Baby_Grouped == "Black" ~ "Black",
      Ethnicity_Baby_Grouped %in% c("Mixed", "Other") ~ "Mixed & Other"
    ),
    Ethnicity_Broad = factor(Ethnicity_Broad,
                              levels = c("White", "Asian", "Black", "Mixed & Other")),
    Time_Binned = ceiling(Time_To_NCC_Hours)
  )

km_fit_broad <- survfit(
  Surv(Time_Binned, rep(1L, nrow(km_clean))) ~ Ethnicity_Broad,
  data = km_clean
)

broad_colours <- c("White" = nhs_blue, "Asian" = nhs_green,
                    "Black" = nhs_pink, "Mixed & Other" = nhs_grey)

# Manual KM plot using ggplot for better control
km_df <- broom::tidy(km_fit_broad) %>%
  mutate(strata = gsub("Ethnicity_Broad=", "", strata))

ggplot(km_df, aes(x = time, y = estimate, colour = strata)) +
  geom_step(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = strata),
              alpha = 0.1, colour = NA) +
  scale_colour_manual(values = broad_colours, name = "Ethnicity") +
  scale_fill_manual(values = broad_colours, name = "Ethnicity") +
  scale_x_continuous(breaks = c(0, 1, 6, 12, 24, 48, 72)) +
  labs(
    title = "Time to NCC Admission by Broad Ethnicity Group",
    subtitle = paste0("FY 2023/24 | N = ", comma(nrow(km_clean)),
                      " admitted babies at NNAP units | Lower curve = faster access"),
    x = "Hours from birth",
    y = "Proportion still awaiting NCC admission"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, colour = "grey40"),
    legend.position = "right"
  )
```

The curves are nearly indistinguishable across broad ethnic groups — consistent with equitable timing of NCC admission.

## Asian Subgroup Disaggregation

```{r km_asian, echo=FALSE, fig.height=7, fig.width=10}
km_asian <- surv_data %>%
  filter(event == 1, At_NNAP_Trust == TRUE,
         !is.na(Ethnicity_Baby_Grouped), !is.na(Time_To_NCC_Hours),
         Time_To_NCC_Hours > 0, Time_To_NCC_Hours <= 72,
         Ethnicity_Baby_Grouped %in% c("White British", "Indian", "Pakistani",
                                        "Bangladeshi", "Other Asian", "Chinese")) %>%
  mutate(
    Ethnicity_Asian = factor(
      as.character(Ethnicity_Baby_Grouped),
      levels = c("White British", "Indian", "Pakistani",
                 "Bangladeshi", "Other Asian", "Chinese")
    ),
    Time_Binned = ceiling(Time_To_NCC_Hours)
  )

km_fit_asian <- survfit(
  Surv(Time_Binned, rep(1L, nrow(km_asian))) ~ Ethnicity_Asian,
  data = km_asian
)

asian_colours <- c("White British" = nhs_blue, "Indian" = nhs_green,
                    "Pakistani" = "#006747", "Bangladeshi" = "#00A499",
                    "Other Asian" = "#78BE20", "Chinese" = nhs_orange)

km_asian_df <- broom::tidy(km_fit_asian) %>%
  mutate(strata = gsub("Ethnicity_Asian=", "", strata))

ggplot(km_asian_df, aes(x = time, y = estimate, colour = strata)) +
  geom_step(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = strata),
              alpha = 0.1, colour = NA) +
  scale_colour_manual(values = asian_colours, name = "Ethnicity") +
  scale_fill_manual(values = asian_colours, name = "Ethnicity") +
  scale_x_continuous(breaks = c(0, 1, 6, 12, 24, 48, 72)) +
  labs(
    title = "Time to NCC Admission: Asian Subgroups vs White British",
    subtitle = paste0("FY 2023/24 | N = ", comma(nrow(km_asian)),
                      " admitted babies | Lower curve = faster access"),
    x = "Hours from birth",
    y = "Proportion still awaiting NCC admission"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, colour = "grey40"),
    legend.position = "right"
  )
```

Disaggregating the Asian group shows no meaningful separation between subgroups. Chinese babies show slightly wider confidence intervals due to smaller numbers.

---

# Cox Frailty Model Results

## Model 6: Primary (Gestation + Unit Type)

```{r model6_table, echo=FALSE}
hr6 <- data.frame(
  Term = names(fe6),
  HR = exp(fe6),
  CI_lower = exp(fe6 - 1.96 * se6),
  CI_upper = exp(fe6 + 1.96 * se6),
  p = 2 * pnorm(-abs(fe6 / se6))
) %>%
  mutate(
    Term = gsub("Ethnicity_Baby_Grouped", "", Term),
    Term = gsub("IMD_Quintile", "IMD: ", Term),
    Term = gsub("Delivery_Unit_Type", "Unit: ", Term),
    `HR (95% CI)` = sprintf("%.3f (%.3f–%.3f)", HR, CI_lower, CI_upper),
    `p-value` = case_when(
      p < 0.001 ~ "<0.001",
      p < 0.01 ~ sprintf("%.3f", p),
      TRUE ~ sprintf("%.2f", p)
    ),
    Significance = case_when(p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*", TRUE ~ "")
  ) %>%
  select(Term, `HR (95% CI)`, `p-value`, Significance)

hr6 %>%
  kable(caption = paste0("Model 6: Cox frailty model — admitted babies at NNAP units (N = ",
                          comma(n_m6), "). Reference: White British / Least Deprived / NICU."),
        escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  pack_rows("Clinical", 1, 1) %>%
  pack_rows("Ethnicity", 2, 10) %>%
  pack_rows("Deprivation", 11, 14) %>%
  pack_rows("Unit Type", 15, 16)
```

::: {.callout-note}
## Reading the Table
HR > 1 = faster admission (better). HR < 1 = slower admission (potential delay). The reference categories are White British, IMD Quintile 5 (Least Deprived), and NICU.
:::

## Model 7: Sensitivity (+ Birth Weight + Apgar)

```{r model7_table, echo=FALSE}
hr7 <- data.frame(
  Term = names(fe7),
  HR = exp(fe7),
  CI_lower = exp(fe7 - 1.96 * se7),
  CI_upper = exp(fe7 + 1.96 * se7),
  p = 2 * pnorm(-abs(fe7 / se7))
) %>%
  mutate(
    Term = gsub("Ethnicity_Baby_Grouped", "", Term),
    Term = gsub("IMD_Quintile", "IMD: ", Term),
    Term = gsub("Delivery_Unit_Type", "Unit: ", Term),
    `HR (95% CI)` = sprintf("%.3f (%.3f–%.3f)", HR, CI_lower, CI_upper),
    `p-value` = case_when(
      p < 0.001 ~ "<0.001",
      p < 0.01 ~ sprintf("%.3f", p),
      TRUE ~ sprintf("%.2f", p)
    ),
    Significance = case_when(p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*", TRUE ~ "")
  ) %>%
  select(Term, `HR (95% CI)`, `p-value`, Significance)

hr7 %>%
  kable(caption = paste0("Model 7: Sensitivity model (N = ", comma(n_m7),
                          "). Reference: White British / Least Deprived / NICU."),
        escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  pack_rows("Clinical", 1, 3) %>%
  pack_rows("Ethnicity", 4, 12) %>%
  pack_rows("Deprivation", 13, 16) %>%
  pack_rows("Unit Type", 17, 18)
```

## Forest Plot

```{r forest_plot, echo=FALSE, fig.height=8, fig.width=10}
forest_data <- data.frame(
  term = names(fe6),
  HR = exp(fe6),
  CI_lower = exp(fe6 - 1.96 * se6),
  CI_upper = exp(fe6 + 1.96 * se6)
) %>%
  filter(!grepl("Gestation_Weeks", term)) %>%
  mutate(
    term = gsub("Ethnicity_Baby_Grouped", "Ethnicity: ", term),
    term = gsub("IMD_Quintile", "IMD: ", term),
    term = gsub("Delivery_Unit_Type", "Unit: ", term),
    category = case_when(
      grepl("Ethnicity", term) ~ "Ethnicity",
      grepl("IMD", term) ~ "Deprivation",
      grepl("Unit", term) ~ "Unit Type"
    ),
    category = factor(category, levels = c("Ethnicity", "Deprivation", "Unit Type")),
    term = factor(term, levels = rev(term))
  )

ggplot(forest_data, aes(x = HR, y = term, colour = category)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper), height = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_colour_manual(values = c(
    "Ethnicity" = nhs_blue, "Deprivation" = nhs_pink, "Unit Type" = nhs_green
  )) +
  facet_grid(category ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Hazard Ratios for Time to NCC Admission (Cox Frailty Model)",
    subtitle = paste0("Admitted babies at NNAP units, FY 2023/24 (N = ", comma(n_m6),
                      "). HR > 1 = faster admission. Ref: White British / Least Deprived / NICU."),
    x = "Hazard Ratio (95% CI)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, colour = "grey40"),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Comparison: Primary vs Sensitivity

```{r comparison_table, echo=FALSE}
# Ethnicity comparison
eth_terms <- grep("Ethnicity", names(fe6), value = TRUE)
imd_terms <- grep("IMD", names(fe6), value = TRUE)
ut_terms <- grep("Delivery_Unit_Type", names(fe6), value = TRUE)

comparison <- tibble(
  Term = c(eth_terms, imd_terms, ut_terms),
  `M6 HR` = exp(fe6[c(eth_terms, imd_terms, ut_terms)]),
  `M7 HR` = exp(fe7[c(eth_terms, imd_terms, ut_terms)]),
) %>%
  mutate(
    Change = `M7 HR` - `M6 HR`,
    Term = gsub("Ethnicity_Baby_Grouped", "", Term),
    Term = gsub("IMD_Quintile", "IMD: ", Term),
    Term = gsub("Delivery_Unit_Type", "Unit: ", Term),
    `M6 HR` = sprintf("%.3f", `M6 HR`),
    `M7 HR` = sprintf("%.3f", `M7 HR`),
    Change = sprintf("%+.3f", Change)
  )

comparison %>%
  kable(caption = "Hazard ratios: Model 6 (primary) vs Model 7 (sensitivity)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

Notable shifts with clinical adjustment:

- Pakistani: HR moves from 0.966 to 1.013 — the apparent 3.4% delay is entirely explained by clinical factors (gestation, birth weight, Apgar)
- Indian: HR moves from 1.008 to 1.046 — slightly *faster* admission after accounting for clinical severity
- LNU effect strengthens: HR 1.028 → 1.037, suggesting the LNU speed advantage is partly independent of casemix

---

# Hospital-Level Variation

```{r frailty, echo=FALSE}
frailty_var_m6 <- unlist(VarCorr(model6))
frailty_var_m7 <- unlist(VarCorr(model7))
```

The shared frailty variance captures between-hospital variation in admission speed, after accounting for patient characteristics and unit type.

| Model | Frailty Variance | Interpretation |
|---|---|---|
| Model 6 (primary) | `r formatC(frailty_var_m6, format = "e", digits = 2)` | Near-zero hospital variation |
| Model 7 (sensitivity) | `r formatC(frailty_var_m7, format = "e", digits = 2)` | Near-zero hospital variation |
| *Phase 1 logistic model* | *ICC ~32%* | *Large hospital variation* |

The contrast with Phase 1 is striking. The logistic model showed that ~32% of variation in *whether* a baby enters NCC is explained by hospital-level factors. Here, the frailty variance is essentially zero — hospitals do not meaningfully differ in *how fast* they admit once the decision is made.

Adding unit type (NICU/LNU/SCBU) absorbed much of the already-small frailty variance from earlier models (which had frailty variance ~0.0005 without unit type). This suggests that what little between-hospital variation existed in admission speed was largely explained by the unit's designation level.

---

# Sensitivity and Diagnostics

## Proportional Hazards Assumption

```{r ph_test, echo=FALSE}
# Fit equivalent coxph for PH diagnostics
ph_data <- surv_data %>%
  filter(event == 1, At_NNAP_Trust == TRUE,
         !is.na(Ethnicity_Baby_Grouped), !is.na(Gestation_Weeks),
         !is.na(IMD_Quintile), !is.na(Delivery_Unit_Type),
         !is.na(Delivery_Site_Code), !is.na(Time_To_NCC_Hours),
         Time_To_NCC_Hours > 0, Time_To_NCC_Hours <= 672) %>%
  mutate(event_admitted = 1L)

model6_coxph <- coxph(
  Surv(Time_To_NCC_Hours, event_admitted) ~
    Gestation_Weeks + Ethnicity_Baby_Grouped + IMD_Quintile + Delivery_Unit_Type,
  data = ph_data,
  ties = "efron"
)

ph_test <- cox.zph(model6_coxph)
```

The Schoenfeld residual test assesses whether hazard ratios are constant over time. With large N, even trivial departures produce small p-values; the correlation coefficient (rho) is more informative — values |rho| > 0.1 suggest practically meaningful departures.

```{r ph_table, echo=FALSE}
ph_df <- data.frame(
  Term = rownames(ph_test$table),
  rho = round(ph_test$table[, 1], 4),
  chisq = round(ph_test$table[, 2], 2),
  p = ph_test$table[, 3]
) %>%
  mutate(
    Term = gsub("Ethnicity_Baby_Grouped", "", Term),
    Term = gsub("IMD_Quintile", "IMD: ", Term),
    Term = gsub("Delivery_Unit_Type", "Unit: ", Term),
    `p-value` = case_when(
      p < 0.001 ~ "<0.001",
      p < 0.01 ~ sprintf("%.3f", p),
      TRUE ~ sprintf("%.2f", p)
    ),
    `|rho|` = abs(rho),
    Concern = ifelse(`|rho|` > 0.1, "Review", "OK")
  ) %>%
  select(Term, rho, chisq, `p-value`, `|rho|`, Concern)

ph_df %>%
  kable(caption = "Proportional hazards test (Schoenfeld residuals)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## NNAP Unit Coverage

The analysis is restricted to births at NHS trusts with a recognised NNAP neonatal unit. This covers `r comma(n_m6)` of `r comma(sum(surv_data$event == 1))` NCC admissions (`r round(100 * n_m6 / sum(surv_data$event == 1), 1)`%). The remaining admissions occurred at delivery sites that could not be linked to an NNAP trust, primarily due to missing delivery site codes in the source data.

```{r nnap_unit_breakdown, echo=FALSE}
unit_breakdown <- admitted_nnap %>%
  filter(!is.na(Delivery_Unit_Type)) %>%
  count(Delivery_Unit_Type, name = "NCC Admissions") %>%
  mutate(`%` = round(100 * `NCC Admissions` / sum(`NCC Admissions`), 1))

unit_breakdown %>%
  kable(caption = "NCC admissions by delivery site unit type (NNAP units only)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

---

# Summary: Findings Across Both Phases

```{r summary_table, echo=FALSE}
summary_df <- tibble(
  Question = c(
    "Who gets admitted to NCC?",
    "How quickly are they admitted?",
    "Does ethnicity affect access?",
    "Does deprivation affect access?",
    "How much do hospitals vary?"
  ),
  `Phase 1 (Logistic)` = c(
    "~13% of births; gestation is the strongest predictor",
    "Not addressed",
    "Modest effects: Pakistani/Chinese lower odds; Black equitable",
    "~8% gradient (most vs least deprived)",
    "ICC ~32% — hospital of birth dominates"
  ),
  `Phase 2 (Survival)` = c(
    "Not addressed (restricted to admitted babies)",
    "Median admission within first few hours; ~70% within 1 hour",
    "No significant delays for any ethnic group after clinical adjustment",
    "No effect — completely flat across IMD quintiles",
    "Near-zero frailty variance — negligible hospital variation in timing"
  )
)

summary_df %>%
  kable(caption = "Comparison of findings across analysis phases") %>%
  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  column_spec(1, bold = TRUE, width = "15em")
```

## Implications

The inequality in NCC access sits at the admission threshold, not the admission process. The intervention point is:

1. Standardising admission criteria across hospitals — the ICC of ~32% in Phase 1 suggests that clinical decision-making about NCC admission varies substantially by hospital. This is where most inequality enters the system.

2. Investigating specific ethnic effects at the hospital level — the modest national-level effects for Pakistani and Chinese babies in Phase 1 may concentrate in specific hospitals or networks.

3. Recognising equitable care delivery — once the admission decision is made, the NHS neonatal system delivers equitable speed-of-access regardless of ethnicity, deprivation, or hospital. This is a positive finding that validates current clinical practice.

::: {.callout-tip}
## Caveat on Timing Resolution
The majority of NCC admissions occur within the first hour of birth. With such a compressed time distribution, the power to detect small ethnic differences in timing is limited. The absence of a statistically significant delay is reassuring but does not rule out very small differences (e.g., minutes) that this analysis cannot resolve.
:::
